{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-09-20 thesis/related.tex
ffinish related

Several other recent advances extend DPOR to new concurrency models,
beyond the shared-memory-threading model outlined in \sect{\ref{sec:landslide-dpor}}.
TransDPOR \cite{transdpor} provides extra domain-specific reduction for message-passing actor programs
by exploiting the fact that the dependency relation is transitive in the absence of shared state.
The $R^4$ algorithm \cite{r4} (corresponding to the R4 checker mentioned above)
extends DPOR to event-driven programs by separating the notion of enabled events from that of multiple threads.
DPOR for TSO and PSO \cite{tsopso}
extends the concurrency model
using {\em shadow threads}, which interleave with traditional threads to represent store buffer nondeterminism,
which can expose bugs not even possible in the strong consistency model
such as discussed in \sect{\ref{sec:tm-warpzone-relaxed}}.
It also introduced a heuristic {\em buffer bounding} technique, analogous to ICB,
to mitigate the corresponding increase in state space size.
The same year, Nidhugg \cite{nidhugg} proposed a DPOR extension to account for TSO and PSO
using {\em chronological traces}.
MCR was recently extended to support relaxed memory models likewise \cite{mcr-tsopso}.
Just this year, RCMC \cite{rcmc} proposed to replace the interleaving model entirely with {\em execution graphs},
which precisely model the executions legal under the RC11 memory model,
offering further reduction still.
Somewhat analogously for HTM, this work's Chapter~\ref{chap:tm}
extended DPOR's concurrency model to include failure injection,
% not "new" -- abstraction reduction is the same as always
and proposed three reduction strategies, one sound and two heuristic,
%to make some progress up the exponential mountain.
to keep state spaces manageable.

@@

It should come as no surprise that many avenues of research towards writing correct programs
have been explored
apart from just executing them a bunch of times to check all the thread interleavings.
Though not as directly related to Landslide as the works referenced above,
this section explores such other approaches,
ranging from building safety guarantees into a language's type system
to checking, proving, and/or enforcing execution properties post-hoc.

@@

While C's extremely rudimentary type system allows the compiler to statically check programs
for properties such as not accidentally dereferencing raw integer values as if they were pointers,
more advanced programming languages may make guarantees about concurrent execution.
%
Erlang \cite{erlang}, an early concurrent functional language,
introduced the actor model for concurrency,
in which threads share no state and must communicate only by message-passing.
%One might say this statically guarantees freedom from data races;
%from another point of view, data races do not even exist as a concept in this programming model.
While this statically guarantees the absence of data races,
programs may still execute nondeterministically,
so concurrency bugs, especially deadlocks, are not ruled out.
Concuerror \cite{concuerror}, discussed above, is a SCT tool for Erlang programs.
%
Haskell \cite{haskell} offers a more sophisticated interface to concurrency:
threads may reference the same objects and even update shared references
using monads that encapsulate mutation,
but at the (garbage-collected) execution level all data is immutable once created,
which preserves type soundness and data-race freedom.
The aforementioned D\'{e}j\`{a} Fu \cite{dejafu} checks concurrent Haskell programs.
%
Rust \cite{rust-language}
presents a type system with more explicit memory management,
in-place mutation,
and mutable references
to appear familiar and approachable to those already versed in C++.
It proposes a borrow-check analysis to ensure memory and type safety
despite mutable references,
and a trait system to ensure no shared state between threads by default.
Its concurrency libraries then offer interfaces which relax this restriction,
allowing threads even to simultaneously reference shared mutable state,
using the type system to enforce sound locking discipline across such accesses,
again preserving type soundness and data-race freedom%
\footnote{The author contributed the original design for this latter feature.}.
I know of no existing SCT for Rust as of yet.
%
The Relaxed Memory Calculus \cite{sully-thesis}
proposes to extend C++ with annotations for weak memory atomics,
which allows for static formal analysis of memory access reorderings.
Although not ruling out data races,
this approach is an important step towards compilers
which can statically reason about program execution under more advanced concurrency models.
%
Finally, LVish \cite{lvish}
features a type system that enforces deterministic behaviour by construction,
using shared state called LVars which allow writes only in ways that update order is not observable.
This renders thread interleavings entirely irrelevant,
obviating any need for runtime verification,
but at the cost of a more restrictive programming model.

@@

Analogous to SCT,
which seeks good coverage of possible thread execution paths under schedule nondeterminism,
another popular testing approach is symbolic execution \cite{symbolic-execution},
which seeks good coverage of possible flow control paths under input nondeterminism.
Symbolic executors
abstract a program's variables and use constraint solving engines such as Z3 \cite{z3}
to work backwards and, effectively,
synthesize combinations of test inputs which can lead to a failure.
KLEE \cite{klee}, one well-known and open-source implementation of this technique,
offers over 90\% code coverage on average across many tests,
often outdoing that achieved by programmers' own hand-written tests.
Later, Contessa \cite{contessa} extended symbolic execution to include concurrency nondeterminism as well,
by using a DPOR-like analysis on individual execution traces
then incorporating reordering possibilities into its SMT constraints.
%rather than using explicit scheduling as SCT does.
This simultaneously exercises both input and schedule nondeterminism,
and achieves good bug-finding performance in practice,
but does not provide the same verification guarantees as repeated DPOR iterations with explicit scheduling.
Exploring both kinds of state space at once thoroughly enough to provide strong verification
is undoubtedly subject to further state space explosion,
and remains future work.

>>>>

Several other recent advances extend DPOR to new concurrency models,
beyond the shared-memory-threading model outlined in \sect{\ref{sec:landslide-dpor}}.
TransDPOR \cite{transdpor} provides extra domain-specific reduction for message-passing actor programs
by exploiting the fact that the dependency relation is transitive in the absence of shared state.
The $R^4$ algorithm \cite{r4} (corresponding to the R4 checker mentioned above)
extends DPOR to event-driven programs by separating the notion of enabled events from that of multiple threads.
TaxDC \cite{taxdc}, a taxonomy study of distributed systems concurrency bugs,
showed that for completeness distributed model checkers must incorporate many forms of nondeterminism,
including message reordering, timeouts, network disconnections, and crashes and reboots, in addition to local threads.
DPOR for TSO and PSO \cite{tsopso}
extends the concurrency model
using {\em shadow threads}, which interleave with traditional threads to represent store buffer nondeterminism,
which can expose bugs not even possible in the strong consistency model
such as discussed in \sect{\ref{sec:tm-warpzone-relaxed}}.
It also introduced a heuristic {\em buffer bounding} technique, analogous to ICB,
to mitigate the corresponding increase in state space size.
The same year, Nidhugg \cite{nidhugg} proposed a DPOR extension to account for TSO and PSO
using {\em chronological traces}.
MCR was recently extended to support relaxed memory models likewise \cite{mcr-tsopso}.
Just this year, RCMC \cite{rcmc} proposed to replace the interleaving model entirely with {\em execution graphs},
which precisely model the executions legal under the RC11 memory model,
offering further reduction still.
Somewhat analogously for HTM, this work's Chapter~\ref{chap:tm}
extended DPOR's concurrency model to include failure injection,
% not "new" -- abstraction reduction is the same as always
and proposed three reduction strategies, one sound and two heuristic,
%to make some progress up the exponential mountain.
to keep state spaces manageable.

@@

%It should come as no surprise that
Naturally, many avenues of research towards writing correct programs
have been explored
apart from just executing them a bunch of times to check all the interleavings.
Though not as directly related as the works referenced above,
this section explores such approaches,
ranging from expressing safety guarantees in a language's type system
to checking, proving, and/or enforcing execution properties post-hoc.

@@

While C's extremely rudimentary type system allows the compiler to statically check programs
for properties such as not accidentally dereferencing raw integer values as if they were pointers,
more advanced programming languages may make guarantees about concurrent execution.
%
Erlang \cite{erlang}, an early concurrent functional language,
introduced the actor model for concurrency,
in which threads share no state and must communicate only by message-passing.
%One might say this statically guarantees freedom from data races;
%from another point of view, data races do not even exist as a concept in this programming model.
While this statically guarantees the absence of data races,
programs may still execute nondeterministically,
so concurrency bugs, especially deadlocks, are not ruled out.
Concuerror \cite{concuerror}, discussed above, is a SCT tool for Erlang programs.
%
Haskell \cite{haskell} offers a more sophisticated interface to concurrency:
threads may reference the same objects and even update shared references
using monads that encapsulate mutation,
but at the (garbage-collected) execution level all data is immutable once created,
which preserves type soundness and data-race freedom.
The aforementioned D\'{e}j\`{a} Fu \cite{dejafu} checks concurrent Haskell programs.
%
Rust \cite{rust-language}
presents a type system with more explicit memory management,
in-place mutation,
and mutable references
to appear familiar and approachable to those already versed in C++.
It proposes a borrow-check analysis to ensure memory and type safety
despite mutable references,
and a trait system to ensure no shared state between threads by default.
Its concurrency libraries then offer interfaces which relax this restriction,
allowing threads even to simultaneously reference shared mutable state,
using the type system to enforce sound locking discipline across such accesses,
again preserving type soundness and data-race freedom%
\footnote{The author themself contributed the original design for this latter feature.}.
I know of no existing SCT for Rust as of yet.
%
The Relaxed Memory Calculus \cite{sully-thesis}
proposes to extend C++ with annotations for weak memory atomics,
which allows for static formal analysis of memory access reorderings.
Although not ruling out data races,
this approach is an important step towards compilers
which can statically reason about program execution under more advanced concurrency models.
%
Finally, LVish \cite{lvish}
features a type system that enforces deterministic behaviour by construction,
using shared state called LVars which allow writes only in ways that update order is not observable.
This renders thread interleavings entirely irrelevant,
obviating any need for runtime verification,
but at the cost of a more restrictive programming model.

@@

Analogous to SCT,
which seeks good coverage of possible thread execution paths under schedule nondeterminism,
another popular testing approach is symbolic execution \cite{symbolic-execution},
which seeks good coverage of possible flow control paths under input nondeterminism.
Symbolic executors
abstract a program's variables and use constraint solvers such as Z3 \cite{z3}
to work backwards and synthesize combinations of test inputs which can lead to a failure.
%
KLEE \cite{klee}, one well-known and open-source implementation,
offers over 90\% code coverage on average across many tests,
often outdoing that achieved by programmers' own hand-written tests.
%
Later, Contessa \cite{contessa} extended symbolic execution to include concurrency nondeterminism as well,
by using a DPOR-like analysis on individual execution traces
then including reordering possibilities in its SMT constraints.
%rather than using explicit scheduling as SCT does.
This simultaneously exercises both input and schedule nondeterminism,
%and achieves good bug-finding performance in practice,
but does not provide the same verification guarantees as repeated DPOR iterations with explicit scheduling.
Exploring both kinds of state space at once thoroughly enough to provide strong verification
is undoubtedly subject to further state space explosion,
and remains future work.
%
Symbiosis \cite{symbiosis} starts from the known root cause of an existing failure
and uses symbolic execution to synthesize a schedule to reproduce it,
then further searches for a non-failing schedule and compares them to produce
a minimum sequence of events necessary for the failure.
This approach skips the initial verification step entirely,
but greatly reduces the diagnosis effort required of the user,
which was a common complaint about Landslide's preemption traces.
